---
title: "Time Series Notes"
output: pdf_document
---

The term "time series" is used to refer to time series data sets and to the processes by which time series data is assumed to be produced.

Time series modeling begins with time series data. Models are built to  

* Describe the data  
* Adjust the data, e.g., remove seasonal components, so that the the long run characteristics of the data can be analyzed and interpreted  
* Make forecasts for the series  
* understand the underlying process which is producing the data.  


##Time series processes

$\{X_t\} t=0, \pm1, ...$ where  $\mu_x(t)=E(X_t)$ and $\gamma_x(r,s)=Cov(X_r, X_s)=E[(X_r-\mu_x(r))(X_s-\mu_x(s))])$.


The underlying assumption of the analysis is that there is a time-independent process at work that is producing the data.  

In other words(Strict Stationarity): $(X_1, ..., X_n) \sim (X_{1+h}, ... , X_{n+h})$


In practice, we focus on the first two moments and define a weaker form of stationarity for $\{X_t\}$.

$\mu_x(t) \amalg t$  and $\gamma_x(t+h, t) \amalg t$ for each $h$

so we simplify to show covariance as a function of lag:  $\gamma_x(h)=Cov(X_t+h, X_t)$

and normalize with the variance $\gamma_x(0)$ to produce the ACF

$$\rho_x(h) \equiv \frac{\gamma_x(h)}{\gamma_x(0)}$$





iid noise
White Noide
Random Walk


MA(1)

$X_t = Z_t + \theta Z_{t-1}, t=0, \pm1, ...$

where $\{Z_t\}~WN(0,\sigma^2)$ and $\theta$ is real.

$$E\left [ X_t \right ]=0$$

$$E\left [ X_{t}^{2} \right ]=\sigma^2(1+\theta^2))$$


$$\gamma_x(t+h, t)=\left\{\begin{matrix}
\sigma^2 (1+\theta^2) & \mathrm{if}\  h=0 \\ 
\sigma^2 \theta & \mathrm{if}\ h= \pm1 \\ 
0 & \mathrm{if}\ \left | h \right |>1
\end{matrix}\right.$$

$$\rho_x(h)=\left\{\begin{matrix}
1 & \mathrm{if}\  h=0 \\ 
\theta / (1+\theta^2) & \mathrm{if}\ h= \pm1 \\ 
0 & \mathrm{if}\ \left | h \right |>1
\end{matrix}\right.$$

```{r}
theta <- 1
z <- rnorm(100)
z.t1<-z[-1]
z<-z[1:length(z.t1)]
x<-z+theta*z.t1
plot(x,main="MA(1) Process")
qqnorm(x)
acf(x)
```


AR(1)



$X_t= \phi X_{t-1}+Z_t$

$\left | \phi  \right |<1$


$\rho_x(h)=\frac{\gamma_x(h)}{\gamma_x(0)}=\phi^{\left | h \right |},  h=0, \pm1, ...$


```{r}
phi <- .8

x[1:100] <- 0

for(i in 2:100){x[i]=phi*x[i-1]+rnorm(1)}

plot(x)
qqnorm(x)
acf(x)
```









##A First approach to time series modeling

One intuitively appealing approach begins by simply observing a time series to be modeled.  For example, look at Australian Wine Sales.

```{r echo=FALSE}
library(itsmr)
wine <- scan("wine.dat")
wine <- as.ts(wine)

yr <- seq(as.Date("1980/1/1"),by = "months",length=length(wine))
plot(yr,wine, type="l",pch=20,main = "Australian Red Wine" , ylab="liters",xlab="Years" )
```

In this plot, the trend and seasonality in the data are clear.  

Out initial approach is view the time series as the sum of three additive components:

$\mathbf{series}_t = \mathbf{trend}_t + \mathbf{seasonality}_t + \mathbf{noise}_t$

in addition, however, you also see A plot of the data shows a clear trend.


trend
seasonality

Sometimes you need to transform a time series

identify and structural components -- trend and seasonality

substracting these components from the series

leaving noise which is then modeled.



```{r}
### transformation
wine.t <- log(wine)
```

```{r echo=FALSE}
# par(mfcol=c(1,2),cex.main=.5,lheight=.5)
# plot(yr,wine, type="l",pch=20,main = "Australian Red Wine (log transformed" , ylab="log(liters)" ,xlab="Years")

plot(yr,wine.t, type="l",pch=20,main = "Australian Red Wine (log transformed" , ylab="feet", xlab="Years" )

```








Lake Huron Data
```{r}
lake <- scan("lake.dat")
lake <- as.ts(lake)
yr <- seq(as.Date("1875/1/1"),by = "months",length=length(lake))
plot(yr,lake, type="l",pch=20,main = "Lake Huron Water Level" , ylab="feet" )
acf=acf(lake)
```

First build a simple linear model to remove the trend.

```{r warning=FALSE, echo=FALSE}
lake1 <- lake[-1]

m <- lake[1:(length(lake1))]

out <- lm(lake1 ~ m)
print(out)
resid <- resid(out)
plot(resid)
acf(resid)
```


##Strikes Series

```{r}
strikes <- scan("strikes.dat")
strikes <- as.ts(strikes)
# use class and structure to tell them apart
yr <- seq(as.Date("1951/1/1"),by = "years",length=length(strikes))
plot(yr,strikes, type="l",pch=20,main = "Strikes in the U.S.A., 1951 -1980" )
acf=acf(strikes)
```




```{r}

rr <- smooth.ma(strikes,2)

plot(yr,strikes, type="l",pch=20,main = "Strikes in the U.S.A., 1951 -1980" )
plot(yr,rr, type="l",pch=20,main = "Strikes in the U.S.A., 1951 -1980" )
rr1 <- strikes - rr
plot(yr,rr1, type="l",pch=20,main = "Strikes in the U.S.A., 1951 -1980" )
acf(rr1)
```








